Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-06-05 article.tex
add todo
Wording about the stacking

\section{Network inference methodology}

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

@@

\section{Signal filtering methodology}

% Explain why we need to filter
% TODO

% Filtering methodology
Prior inferring the network, the fluorescent calcium time series are smoothed
using several linear and non linear filters. A low-pass filter, a high-pass
filter and a non linear hard-thresholding based filter are
applied for each possible time step $t$
\begin{align}
X^\prime_{t,i} &= \sum_{l=-3}^2 c_l X_{t+l,i} \forall i, \\
X^{\prime\prime}_{t,i} &= X^{\prime}_{t,i} - X^{\prime}_{t-1,i} \forall i, \\
X^{\prime\prime\prime}_{t,i} &= {X^{\prime\prime}_{t,i}}^{0.9} I(X^{\prime\prime}_{t,i} \geq f_3) \forall i,
\end{align}
where the $c_l$ coefficients were taken in one of the four following tuples
$(\left\{c_l\right\}_{l=-3}^2)=(0, 0, 1, 1, 1, 1)$,
$(\left\{c_l\right\}_{l=-3}^2)=(0, 0, 1, 1, 1, 0)$,
$(\left\{c_l\right\}_{l=-3}^2)=(1, 1, 1, 1, 0, 0)$ or
$(\left\{c_l\right\}_{l=-3}^2)=(0.4, 0.8, 1, 1, 0, 0)$ and $f_3$ is a parameter
of the hard-threshold filter.

Then we apply a non-linear filter based on the current
activity of the systems in order to favor the association measure when
a medium size set of neurons are firing together, while artificially
reducing the association measure in case of extreme network activity,
i.e. high when a network firing is happening and low when no neuron is
active:
\begin{align}
a_t &= \sum_{i=1}^p X^{\prime\prime\prime}_{t,i} +
                   \frac{1}{2} X^{\prime\prime\prime}_{t-1,i} \forall t \in T\\
r_t &= \frac{a_t}{\max_{l \in T}{a_l}} \\
X^{\prime\prime\prime\prime}_{t,i} &=
\left\{
  \begin{array}{l}
    1  : r_t < 0\\
    {(X^{\prime\prime\prime}_{t,i} + 1)^{(a_t^{-1} + 1)}}^{g(r_t)} : r_t \geq 0
  \end{array}
\right.
\end{align}
where $g(x) = b_1 I(f_4 < x < f_5) + b_2  I(f_5 \leq x < f_6) +
b_3 (I(x\leq f_4) + I(f_ 6 \leq x))$ is a piecewise continuous function of
parameters $b_1, b_2, b_3, f_4, f_5, f_6$. The coefficient were tuned
for each set of coefficient $\left\{c_l\right\}_{l=-3}^2$.

\section{Stacking methodology}

% Speak about why averaging
In order to improve performance, we stacked the partial correlation matrix
from a reasonable of combinaison TODO continue

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
Since the partial correlation is a symmetric, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both ) can't
be inferred.\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\cite{de2004discovery}.} \cite{de2004discovery}.
In our last submission in hope to improve performance, some causal information
(directivity of the links) were added by stacking the partial correlation score
matrix with an pairwise activation heuristics. This method computes an anti-symmetric
activation score $z_{ij}$ based the variation of fluorescence signal of a neuron
$j$ due to a neuron $i$ over all timesteps:
\[
z_{ij} = \sum_{t=1}^n I(X_{t+1,j} - X_{t, i} \in [f_1, f_2]) -  I(X_{t+1,i} - X_{t, j} \in [f_1, f_2])
\]
where $f_1$ and $f_2$ are parameters of the method and $I$ is the
indicator function.

>>>>

\section{Signal filtering methodology}
\label{sec:filter}
% Explain why we need to filter
% TODO

% Filtering methodology
Prior inferring the network, the fluorescent calcium time series are smoothed
using several linear and non linear filters. A low-pass filter, a high-pass
filter and a non linear hard-thresholding based filter are
applied for each possible time step $t$
\begin{align}
X^\prime_{t,i} &= \sum_{l=-3}^2 c_l^k X_{t+l,i} \forall i, \\
X^{\prime\prime}_{t,i} &= X^{\prime}_{t,i} - X^{\prime}_{t-1,i} \forall i, \\
X^{\prime\prime\prime}_{t,i} &= {X^{\prime\prime}_{t,i}}^{0.9} I(X^{\prime\prime}_{t,i} \geq f_3) \forall i,
\end{align}
where the $c_l^k$ coefficients were taken in one of the four following tuples
$\left\{c_l^1\right\}=(0, 0, 1, 1, 1, 1)$,
$\left\{c_l^2\right\}=(0, 0, 1, 1, 1, 0)$,
$\left\{c_l^3\right\}=(1, 1, 1, 1, 0, 0)$ or
$\left\{c_l^4\right\}=(0.4, 0.8, 1, 1, 0, 0)$ and $f_3$ is a parameter
of the hard-threshold filter.

Then we apply a non-linear filter based on the current
activity of the systems in order to favor the association measure when
a medium size set of neurons are firing together, while artificially
reducing the association measure in case of extreme network activity,
i.e. high when a network firing is happening and low when no neuron is
active:
\begin{align}
a_t &= \sum_{i=1}^p X^{\prime\prime\prime}_{t,i} +
                   \frac{1}{2} X^{\prime\prime\prime}_{t-1,i} \forall t \in T\\
r_t &= \frac{a_t}{\max_{l \in T}{a_l}} \\
X^{\prime\prime\prime\prime}_{t,i} &=
\left\{
  \begin{array}{l}
    1  : r_t < 0\\
    {(X^{\prime\prime\prime}_{t,i} + 1)^{(a_t^{-1} + 1)}}^{g(r_t)} : r_t \geq 0
  \end{array}
\right.
\end{align}
where $g(x) = b_1 I(f_4 < x < f_5) + b_2  I(f_5 \leq x < f_6) +
b_3 (I(x\leq f_4) + I(f_ 6 \leq x))$ is a piecewise continuous function of
parameters $b_1, b_2, b_3, f_4, f_5, f_6$. The coefficient were tuned
for each set of coefficient $\left\{c_l^k\right\}$.

\section{Network inference methodology}
\label{sec:inference}
% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

@@

\section{Stacking methodology}
\label{sec:stacking}
% Speak about why averaging
%In order to improve performance, we stacked the partial correlation matrix

Our first guess was that introduce an ensemble method would improve performance
[cite papers which praise ensemble methods]. Among all parameters of our method,
we choose to average over two parameters, i.e. over multiple combinations of the
hard-threshold value $f_3$ and the pattern of the low-pass filter $c_l$. In order to 
free ourself from the choice of a probability distribution for the $(f_3,c_l)$ combinations,
we choose a uniform distribution for the threshold distribution (in a given interval of 
values) within each filtering process and perform a weighted average for the low-pass 
filter pattern. Formally, the low-pass filter

\begin{align}
y_{stacking} = \sum_{k=1}^{k=4}  w_k \sum_{f_3 \in T} f(X, k, f_3)
\end{align}
where $f(X, k, f_3)$ applies the filtering and inference methodologies (as described 
in sections \ref{sec:filter} and \ref{sec:inference}) with a  given value $f_3$ for the
hard-threshold among a set of thresholds $T$ and the $k^{th}$ low-pass filter pattern.

\section{Causal discovery methodology}
% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
Since the partial correlation is a symmetric, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both ) can't
be inferred\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\cite{de2004discovery}.}\cite{de2004discovery}.
In our last submission in hope to improve performance, some causal information
(directivity of the links) were added by stacking the partial correlation score
matrix with an pairwise activation heuristics. This method computes an anti-symmetric
activation score $z_{ij}$ based the variation of fluorescence signal of a neuron
$j$ due to a neuron $i$ over all timesteps:
\[
z_{ij} = \sum_{t=1}^n I(X_{t+1,j} - X_{t, i} \in [f_1, f_2]) -  I(X_{t+1,i} - X_{t, j} \in [f_1, f_2])
\]
where $f_1$ and $f_2$ are parameters of the method and $I$ is the
indicator function. % TODO say that we use only the 3 first filters for this
